---
title: "2_25"
author: "Aarushi Sahejpal"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Lesson**

```{r}

# ---- Step 1: Load necessary libraries ----
# Install packages if you don't have them already
# install.packages(c("tidyverse", "lubridate", "scales", "knitr"))

library(tidyverse)     # For data manipulation and visualization
library(lubridate)     # For handling dates
library(scales)        # For better axis formatting
library(knitr)         # For nice tables

# ---- Step 2: Import the data ----
# Set your working directory to where the file is located
# setwd("your/path/here")

# Import the CSV file
violations <- read_csv("landlord_violations.csv")
```

```{r}
# group by 
# count
# Median and mean
# more aggregates
```


**Explore first**

```{r}


# ---- Step 3: Basic data exploration ----
# Look at the dimensions of the dataset
summary(violations) # Shows number of rows and columns

# Look at the structure of the data
str(violations)

# View the first few rows
head(violations)

# View the last few rows
tail(violations)

```

```{r}

# Get column names
colnames(violations)

# Count total number of rows
cat("Total number of violations:", nrow(violations), "\n")

# Count unique values in key columns
cat("Number of unique landlords:", n_distinct(violations$OWNER, na.rm = TRUE), "\n")
cat("Number of unique addresses:", n_distinct(violations$`VIOLATION ADDRESS`, na.rm = TRUE), "\n")
cat("Number of unique violation types:", n_distinct(violations$VIOLATION, na.rm = TRUE), "\n")

# Basic counts by category
cat("\nViolations by floor:\n")
print(table(violations$FLOOR))

cat("\nViolations by location:\n")
head(sort(table(violations$LOCATION), decreasing = TRUE), 10)

# Check for missing values in each column
cat("\nNumber of missing values by column:\n")
colSums(is.na(violations))

# Get a sense of the time range
if(is.character(violations$`CREATED DATE`)) {
  # Convert to Date format for this calculation
  date_col <- as.Date(violations$`CREATED DATE`, format="%m/%d/%Y")
  cat("\nDate range of violations:", min(date_col, na.rm = TRUE), "to", max(date_col, na.rm = TRUE), "\n")
}

# Initial exploration of fine amounts
if(grepl("\\$", violations$`ORIGINAL FINE AMOUNT`[1])) {
  # Convert to numeric for this calculation
  fine_col <- as.numeric(gsub("\\$|,", "", violations$`ORIGINAL FINE AMOUNT`))
  
  cat("\nFine amount summary:\n")
  cat("Min:", min(fine_col, na.rm = TRUE), "\n")
  cat("Max:", max(fine_col, na.rm = TRUE), "\n")
  cat("Mean:", mean(fine_col, na.rm = TRUE), "\n")
  cat("Median:", median(fine_col, na.rm = TRUE), "\n")
  
  # Create a simple histogram of fine amounts
  hist(fine_col, main = "Distribution of Fine Amounts", 
       xlab = "Fine Amount ($)", col = "lightblue", breaks = 20)
}

# Group by OWNER and summarize
owner_summary <- violations %>%
  group_by(OWNER) %>%
  summarise(
    count = n(),
    avg_fine = mean(as.numeric(gsub("\\$|,", "", `ORIGINAL FINE AMOUNT`)), na.rm = TRUE),
    min_fine = min(as.numeric(gsub("\\$|,", "", `ORIGINAL FINE AMOUNT`)), na.rm = TRUE),
    max_fine = max(as.numeric(gsub("\\$|,", "", `ORIGINAL FINE AMOUNT`)), na.rm = TRUE)
  ) %>%
  arrange(desc(count))

# Print top landlords by violation count
cat("\nTop 10 landlords by violation count:\n")
print(head(owner_summary, 10))

# Group by location and summarize
location_summary <- violations %>%
  group_by(LOCATION) %>%
  summarise(
    count = n(),
    avg_fine = mean(as.numeric(gsub("\\$|,", "", `ORIGINAL FINE AMOUNT`)), na.rm = TRUE)
  ) %>%
  arrange(desc(count))

# Print top locations by violation count
cat("\nTop 10 locations by violation count:\n")
print(head(location_summary, 10))

# ---- Step 4: Data cleaning and preparation ----
# Convert date column to Date type
violations <- violations %>%
  mutate(CREATED_DATE = mdy(`CREATED DATE`))

# Convert fine amount to numeric
violations <- violations %>%
  mutate(FINE_AMOUNT = as.numeric(gsub("\\$|,", "", `ORIGINAL FINE AMOUNT`)))

# Look at missing values
missing_values <- violations %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), 
               names_to = "Column", 
               values_to = "Missing_Count")

print(missing_values)

# ---- Step 5: Basic exploratory analysis ----
# 1. Distribution of violations by location
location_counts <- violations %>%
  count(LOCATION, sort = TRUE) %>%
  top_n(10)

# Create a bar chart for top 10 locations
ggplot(location_counts, aes(x = reorder(LOCATION, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Violation Locations",
       x = "Location",
       y = "Number of Violations") +
  theme_minimal()

# 2. Distribution of fine amounts
ggplot(violations, aes(x = FINE_AMOUNT)) +
  geom_histogram(bins = 20, fill = "coral") +
  scale_x_continuous(labels = dollar_format()) +
  labs(title = "Distribution of Fine Amounts",
       x = "Fine Amount ($)",
       y = "Count") +
  theme_minimal()

# 3. Violations over time
violations_by_date <- violations %>%
  count(CREATED_DATE) 

ggplot(violations_by_date, aes(x = CREATED_DATE, y = n)) +
  geom_line() +
  labs(title = "Number of Violations Over Time",
       x = "Date",
       y = "Number of Violations") +
  theme_minimal()

# ---- Step 6: Analyze top landlords with violations ----
top_landlords <- violations %>%
  count(OWNER, sort = TRUE) %>%
  filter(!is.na(OWNER)) %>%
  top_n(10)

ggplot(top_landlords, aes(x = reorder(OWNER, n), y = n)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  coord_flip() +
  labs(title = "Top 10 Landlords by Number of Violations",
       x = "Landlord",
       y = "Number of Violations") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))  # Make text smaller to fit

# ---- Step 7: Analyze violation types ----
# Extract the main violation type from the detailed violation text
violations <- violations %>%
  mutate(VIOLATION_TYPE = str_extract(VIOLATION, "^[^|]+"))

# Count the top violation types
top_violations <- violations %>%
  count(VIOLATION_TYPE, sort = TRUE) %>%
  top_n(10)

ggplot(top_violations, aes(x = reorder(VIOLATION_TYPE, n), y = n)) +
  geom_bar(stat = "identity", fill = "purple") +
  coord_flip() +
  labs(title = "Top 10 Violation Types",
       x = "Violation Type",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))  # Make text smaller to fit

# ---- Step 8: Analyze fine amounts by violation type ----
fine_by_violation <- violations %>%
  group_by(VIOLATION_TYPE) %>%
  summarise(
    avg_fine = mean(FINE_AMOUNT, na.rm = TRUE),
    total_fine = sum(FINE_AMOUNT, na.rm = TRUE),
    count = n()
  ) %>%
  arrange(desc(avg_fine)) %>%
  filter(count > 10)  # Only include violation types with more than 10 occurrences

# Show top 10 violations by average fine
top_10_by_fine <- head(fine_by_violation, 10)
kable(top_10_by_fine)

# Create a scatter plot showing relationship between violation count and average fine
ggplot(fine_by_violation, aes(x = count, y = avg_fine)) +
  geom_point(aes(size = total_fine), alpha = 0.7, color = "blue") +
  scale_y_continuous(labels = dollar_format()) +
  scale_size_continuous(name = "Total Fines", labels = dollar_format()) +
  labs(title = "Violation Count vs. Average Fine Amount",
       x = "Number of Violations",
       y = "Average Fine Amount ($)") +
  theme_minimal()

# ---- Step 9: Geographic analysis ----
# Extract street names from violation addresses
violations <- violations %>%
  mutate(STREET = str_extract(`VIOLATION ADDRESS`, "[A-Z]+ (ST|AVE|BLVD|RD|DR|LN|CT|PL|WAY|TER|CIR)"))

# Count violations by street
street_counts <- violations %>%
  count(STREET, sort = TRUE) %>%
  filter(!is.na(STREET)) %>%
  top_n(15)

ggplot(street_counts, aes(x = reorder(STREET, n), y = n)) +
  geom_bar(stat = "identity", fill = "orange") +
  coord_flip() +
  labs(title = "Top 15 Streets with Most Violations",
       x = "Street",
       y = "Number of Violations") +
  theme_minimal()


# ---- Step 12: Seasonal patterns in violations ----
violations <- violations %>%
  mutate(month = month(CREATED_DATE),
         month_name = month(CREATED_DATE, label = TRUE))

monthly_counts <- violations %>%
  count(month, month_name)

ggplot(monthly_counts, aes(x = month_name, y = n, group = 1)) +
  geom_line() +
  geom_point() +
  labs(title = "Violations by Month",
       x = "Month",
       y = "Number of Violations") +
  theme_minimal()

# ---- Step 13: Create a summary report ----
# Calculate key statistics
total_violations <- nrow(violations)
total_landlords <- n_distinct(violations$OWNER, na.rm = TRUE)
total_addresses <- n_distinct(violations$`VIOLATION ADDRESS`, na.rm = TRUE)
avg_fine <- mean(violations$FINE_AMOUNT, na.rm = TRUE)
max_fine <- max(violations$FINE_AMOUNT, na.rm = TRUE)

# Print summary statistics
cat("Summary of DC Landlord Violations:\n")
cat("Total Violations:", total_violations, "\n")
cat("Total Unique Landlords:", total_landlords, "\n")
cat("Total Unique Addresses:", total_addresses, "\n")
cat("Average Fine Amount: $", round(avg_fine, 2), "\n")
cat("Maximum Fine Amount: $", max_fine, "\n")

```

**Re-create story findings.**

1.  As of Aug. 1, 2024, the owners of 3003 Van Ness St. NW owed the District more than `$1.2 million` in unpaid fees for more than 700 violations since 2018, the most of any residential building in D.C during that time frame. That amount dropped significantly after reporters began asking questions in recent months, and as of Jan. 15, 2025, the amount of housing code fines totaled `$182,000` for 139 violations.
2.  The remaining `$182,000` in unpaid fines at the Van Ness property is just a sliver of the more than `$40 million` in unpaid fees amassed by landlords across the District since 2018, according to a new analysis conducted by the *Investigative Reporting Workshop*.


**Create a bar chart of top landlords by fine amounts **

```{r}
# Load necessary libraries
library(dplyr)
library(lubridate)

# Read the CSV file into a data frame
violations_data <- read.csv("landlord_violations.csv")

# Convert 'CREATED DATE' column to Date type if it's not already
violations_data$CREATED.DATE <- as.Date(violations_data$CREATED.DATE, format = "%Y-%m-%d")

# Filter the data for violations at '3003 VAN NESS ST NW' between 2018 and 1 Aug 2024
start_date <- as.Date("2018-01-01")
end_date <- as.Date("2024-08-01")

filtered_data <- violations_data %>%
  filter(VIOLATION.ADDRESS == "3003 VAN NESS ST NW" &
           CREATED.DATE >= start_date &
           CREATED.DATE <= end_date)

# Count the number of violations
violation_count <- nrow(filtered_data)

# Output the result
print(paste("Number of violations at 3003 VAN NESS ST NW between 2018 and 1 Aug 2024:", violation_count))

```
Code didn't work as it says 0, trying to check it 
```{r}
head(violations_data$CREATED.DATE)

```
```{r}
head(violations_data$CREATED.DATE)

```
```{r}
violations_data$CREATED.DATE <- as.Date(violations_data$CREATED.DATE, format = "%m/%d/%Y")
```


```{r}
head(violations_data$CREATED.DATE)
```
```{r}
# Find rows where the date conversion failed
invalid_dates <- violations_data[is.na(violations_data$CREATED.DATE), ]
head(invalid_dates)

```
```{r}
# Load necessary libraries
library(dplyr)
library(lubridate)

# Read the CSV file into a data frame
violations_data <- read.csv("1.Detail Landlord Violations.xlsx - Sheet 1.csv")

# Clean up the address column (remove leading/trailing spaces)
violations_data$VIOLATION.ADDRESS <- trimws(violations_data$VIOLATION.ADDRESS)

# Convert 'CREATED DATE' column to Date type if it's not already
violations_data$CREATED.DATE <- as.Date(violations_data$CREATED.DATE, format = "%m/%d/%Y")  # Adjust format if necessary

# Check column names and inspect some data
colnames(violations_data)
head(violations_data$CREATED.DATE)
head(violations_data$VIOLATION.ADDRESS)

# Filter the data for violations at '3003 VAN NESS ST NW' between 2018 and 1 Aug 2024
start_date <- as.Date("2018-01-01")
end_date <- as.Date("2024-08-01")

filtered_data <- violations_data %>%
  filter(VIOLATION.ADDRESS == "3003 VAN NESS ST NW" &
         CREATED.DATE >= start_date &
         CREATED.DATE <= end_date)

# Count the number of violations
violation_count <- nrow(filtered_data)

# Output the result
print(paste("Number of violations at 3003 VAN NESS ST NW between 2018 and 1 Aug 2024:", violation_count))

```
```{r}
# Load necessary libraries
library(dplyr)

# Load the data from the CSV file
data <- read.csv("1.Detail Landlord Violations.csv")

# Count how many times '3003 VAN NESS ST NW' appears in the 'VIOLATION ADDRESS' column
count_violations <- data %>%
  filter(`VIOLATION ADDRESS` == "3003 VAN NESS ST NW") %>%
  nrow()

# Print the result
print(count_violations)

```
```{r}
# Load necessary libraries
library(readxl)
library(dplyr)

# Load the data from the Excel file
data <- read_excel("1.Detail Landlord Violations.xlsx", sheet = "Sheet 1")

# Count how many times '3003 VAN NESS ST NW' appears in the 'VIOLATION ADDRESS' column
count_violations <- data %>%
  filter(`VIOLATION ADDRESS` == "3003 VAN NESS ST NW") %>%
  nrow()

# Print the result
print(count_violations)

```
```{r}
# Load necessary libraries
library(dplyr)

# Load the data from the CSV file using the full path
data <- read.csv("~/Uni/MA Journalism/Learn R/1.Detail Landlord Violations.xlsx - Sheet 1.csv")

# Count how many times '3003 VAN NESS ST NW' appears in the 'VIOLATION ADDRESS' column
count_violations <- data %>%
  filter(`VIOLATION ADDRESS` == "3003 VAN NESS ST NW") %>%
  nrow()

# Print the result
print(count_violations)

```

```{r}
# Load necessary libraries
library(dplyr)

# Load the data from the CSV file
data <- read.csv("~/Uni/MA Journalism/Learn R/1.Detail Landlord Violations.xlsx - Sheet 1.csv")

# Print column names to check
colnames(data)

```
```{r}
# Load necessary libraries
library(dplyr)

# Load the data from the CSV file
data <- read.csv("~/Uni/MA Journalism/Learn R/1.Detail Landlord Violations.xlsx - Sheet 1.csv")

# Count how many times '3003 VAN NESS ST NW' appears in the 'VIOLATION.ADDRESS' column
count_violations <- data %>%
  filter(`VIOLATION.ADDRESS` == "3003 VAN NESS ST NW") %>%
  nrow()

# Print the result
print(count_violations)

```
```{r}
# Load necessary libraries
library(dplyr)

# Load the data from the CSV file
data <- read.csv("~/Uni/MA Journalism/Learn R/1.Detail Landlord Violations.xlsx - Sheet 1.csv")

# Clean up the 'VIOLATION.ADDRESS' column
data$`VIOLATION.ADDRESS` <- trimws(tolower(data$`VIOLATION.ADDRESS`))  # Remove spaces and make lowercase

# Count how many times '3003 VAN NESS ST NW' appears in the 'VIOLATION.ADDRESS' column (case-insensitive)
count_violations <- data %>%
  filter(`VIOLATION.ADDRESS` == tolower(trimws("3003 VAN NESS ST NW"))) %>%
  nrow()

# Print the result
print(count_violations)

```
```{r}
# Load necessary libraries
library(dplyr)

# Load the data from the CSV file
data <- read.csv("~/Uni/MA Journalism/Learn R/1.Detail Landlord Violations.xlsx - Sheet 1.csv")

# Clean up the 'OWNER' column (remove extra spaces and convert to lowercase)
data$OWNER <- trimws(tolower(data$OWNER))

# Count how many times 'SMITH PROPERTY HOLDINGS VAN NESS L.P.' appears in the 'OWNER' column
count_owner <- data %>%
  filter(OWNER == tolower(trimws("SMITH PROPERTY HOLDINGS VAN NESS L.P."))) %>%
  nrow()

# Print the result
print(count_owner)

```
```{r}
# Load necessary libraries
library(dplyr)

# Load the data from the CSV file
data <- read.csv("~/Uni/MA Journalism/Learn R/1.Detail Landlord Violations.xlsx - Sheet 1.csv")

# Clean up the 'VIOLATION.ADDRESS' column (remove spaces and convert to lowercase)
data$`VIOLATION.ADDRESS` <- trimws(tolower(data$`VIOLATION.ADDRESS`))

# Clean up the 'ORIGINAL.FINE.AMOUNT' column (ensure it's numeric, removing any non-numeric characters if needed)
data$`ORIGINAL.FINE.AMOUNT` <- as.numeric(gsub("[^0-9.-]", "", data$`ORIGINAL.FINE.AMOUNT`))

# Calculate the total 'ORIGINAL.FINE.AMOUNT' for '3003 VAN NESS ST NW'
total_fine <- data %>%
  filter(`VIOLATION.ADDRESS` == tolower(trimws("3003 VAN NESS ST NW"))) %>%
  summarise(total_fine = sum(`ORIGINAL.FINE.AMOUNT`, na.rm = TRUE))

# Print the result
print(total_fine)

```
  #Post Aarushi feedback!! 
```{r}
# Load necessary libraries

library(dplyr)

library(lubridate)

 

# Read the CSV file into a data frame

violations_data <- read.csv("~/Uni/MA Journalism/Learn R/1.Detail Landlord Violations.xlsx - Sheet 1.csv")

 

# Convert 'CREATED DATE' column to Date type - fix the format

violations_data$CREATED.DATE <- mdy(violations_data$CREATED.DATE)  # Use mdy() for MM/DD/YYYY format

 

# Filter for all available data for 3003 Van Ness St NW

van_ness_violations <- violations_data %>%

  filter(`VIOLATION.ADDRESS` == "3003 VAN NESS ST NW")

 

# Count the number of violations

violation_count <- nrow(van_ness_violations)

 

# Calculate total fine amount (after converting from string to numeric)

van_ness_violations$FINE_AMOUNT <- as.numeric(gsub("[$,]", "", van_ness_violations$ORIGINAL.FINE.AMOUNT))

total_fines <- sum(van_ness_violations$FINE_AMOUNT, na.rm = TRUE)

 

# Output the result

print(paste("Number of violations at 3003 VAN NESS ST NW in the available data:", violation_count))

print(paste("Total fine amount: $", format(total_fines, big.mark=",")))

 

# For Jan 15, 2025 cutoff

jan_15_violations <- van_ness_violations %>%

  filter(CREATED.DATE <= as.Date("2025-01-15"))

jan_15_count <- nrow(jan_15_violations)

jan_15_fines <- sum(jan_15_violations$FINE_AMOUNT, na.rm = TRUE)

 

print(paste("Violations until Jan 15, 2025:", jan_15_count))

print(paste("Fine amount until Jan 15, 2025: $", format(jan_15_fines, big.mark=",")))

```
```
That did work! Will now try and find total fine amount, equivalent to the $40 million figure in the original story.
```{r}
total_fines <- sum(van_ness_violations$FINE_AMOUNT, na.rm = TRUE)
print(paste("Total fine amount: $", format(total_fines, big.mark = ",")))

```

```{r}
# Clean 'ORIGINAL.FINE.AMOUNT' by removing $ and commas
violations_data$ORIGINAL.FINE.AMOUNT <- as.numeric(gsub("[$,]", "", violations_data$ORIGINAL.FINE.AMOUNT))

# Calculate the total fine amount
total_fines <- sum(violations_data$ORIGINAL.FINE.AMOUNT, na.rm = TRUE)

# Output the result
print(paste("Total fine amount: $", format(total_fines, big.mark = ",")))

```
For some reason it is now not outputting up here as it won't let me press play? But the answer from the console is $8,410,288
```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)

# Clean 'ORIGINAL.FINE.AMOUNT' column by removing $ and commas
violations_data$ORIGINAL.FINE.AMOUNT <- as.numeric(gsub("[$,]", "", violations_data$ORIGINAL.FINE.AMOUNT))

# Summarize the data by OWNER and sum the 'ORIGINAL.FINE.AMOUNT'
owner_fines <- violations_data %>%
  group_by(OWNER) %>%
  summarize(total_fines = sum(ORIGINAL.FINE.AMOUNT, na.rm = TRUE)) %>%
  arrange(desc(total_fines))

# Get the top 10 owners with the highest fines (you can adjust the number of top owners here)
top_owners <- head(owner_fines, 10)

# Create a bar chart using ggplot2
ggplot(top_owners, aes(x = reorder(OWNER, total_fines), y = total_fines, fill = OWNER)) +
  geom_bar(stat = "identity") +
  coord_flip() +  # Flip coordinates to make the bars horizontal
  labs(
    title = "Top 10 Owners by Total Fine Amount",
    x = "Owner",
    y = "Total Fine Amount ($)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels if needed

```
```{r}
# Create a bar chart using ggplot2 with larger bars and no legend
ggplot(top_owners, aes(x = reorder(OWNER, total_fines), y = total_fines)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.8) +  # Increased bar size with width
  coord_flip() +  # Flip coordinates to make the bars horizontal
  labs(
    title = "Top 10 Owners by Total Fine Amount",
    x = "Owner",
    y = "Total Fine Amount ($)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "none",  # Remove the legend
    axis.title.x = element_text(size = 14),  # Increase size of axis titles
    axis.title.y = element_text(size = 14),  # Increase size of axis titles
    plot.title = element_text(size = 16, face = "bold")  # Increase size of plot title
  )

```

Adding labels with the amounts - this is also less readable because the top one I think is the NA column, which if I was writing a story on it I would want to dig further into that data. 
```{r}
# Create a bar chart using ggplot2 with larger bars, no legend, and labels
ggplot(top_owners, aes(x = reorder(OWNER, total_fines), y = total_fines)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.8) +  # Increased bar size
  geom_text(aes(label = scales::dollar(total_fines)), hjust = -0.1, size = 5) +  # Add labels with total fine amount
  coord_flip() +  # Flip coordinates to make the bars horizontal
  labs(
    title = "Top 10 Owners by Total Fine Amount",
    x = "Owner",
    y = "Total Fine Amount ($)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "none",  # Remove the legend
    axis.title.x = element_text(size = 14),  # Increase size of axis titles
    axis.title.y = element_text(size = 14),  # Increase size of axis titles
    plot.title = element_text(size = 16, face = "bold")  # Increase size of plot title
  )

```
